Learning from Data Through Models

### 2023-07-13 paper/paper.Rmd
regressions in r

---
title: "Learning from Data Through Models"
date: '`r format(Sys.time(), "%B %d, %Y")`'
author: "Alberto Bisin, Guillaume Frechette and Jimena Galindo"
abstract: "TBW"
output: 
    pdf_document:
        citation_package: natbib
        fig_width: 7
        fig_height: 6
        fig_caption: true
        number_sections: false
        keep_tex: true
        extra_dependencies:
            footmisc: ["bottom"] # footnote management
            setspace: ["doublespacing"] # spacing of the paper
            caption: ["normal"]
            dsfont: null # indicator function 1
            booktabs: null
            makecell: null
            hyperref: null

bibliography: bibliography.bib
fontsize: 12pt
geometry:
- margin=2.5cm
hyperref:
- linestretch=2.0
documentclass: article
---

```{r dependencies, include=FALSE}
library(data.table)
library(here)
renv::snapshot()
library("languageserver")
library(reticulate)
```

# Introduction
TBW

# Literature Review
TBW

@@

### Optimal Rules
Conditional of using a particular mental model, the most accurate prediction rule that the agent can use is one that 
predicts the state to be **B** whenever, given the observables under consideration, **B** is more likely. 
And it predicts **R** whenever **R** is more likely. Because we abstract away from the procedure through which the rule is updated, 
we will make predictions under the assumption that the agent has learned the parameters perfectly. That is, 
they have seen enough information that, under an updating procedure that converges to the truth---for example, 
correctly specified regression model or Bayes rule---they would have converged to the true parameters already. This provides an 
upper bound for how well any model can perform.

@@

##### Proposition.
_Fix any model that considers $m\leq N$ variables and relabel the variable so that $x_1, ..., x_m$ are the variables 
that the model considers and $x_{m+1}, ..., x_N$ are the ones that the model does not consider. Relabel the 
coefficients $a = (a_1, ..., a_N)$ accordingly as well._

##### proof.
_Let $M:=a_1x_1+...+a_mx_m$ and $M^c = a_{m+1}x_{m+1}+...+a_Nx_N+k$ and define the latent variable $y^L := M +  M^c$. 
The optimal prediction rule for model $M$ predicts the state to be $\textbf{B}$ whenever 
$P[y^L|M]\geq \frac{1}{2}$ and predicts $\textbf{R}$ otherwise._

_Using the fact that for a random variable $z\sim N(\mu_z, \sigma_z)$ we have that $P[z\geq \mu_z]=\frac{1}{2}$ and 
noticing that $M^c$ and $y^L|M$ are Normally distributed with means $\mathbb{E}[M^c]$ and $M+\mathbb{E}[M^c]$ 
respectively, it is easy to see that $P[y^L\geq 0 | M] \geq P[y^L\geq M+\mathbb{E}[M^c | M]]=\frac{1}{2}$ whenever 
$M+\mathbb{E}[M^c]\geq 0$. Similarly, whenever $M+\mathbb{E}[M^c]< 0$ we will have that 
$P[y^L\geq 0 | M] < P[y^L\geq M+\mathbb{E}[M^c | M]]=\frac{1}{2}$. Therefore, the optimal rule is to predict 
$\mathbf{B}$ whenever $a_1x_1+...+a_kx_k\geq -\mathbb{E}[M^c]$ and $\textbf{R}$ otherwise._

@@

### Polarization
We take the definition of polarization from @Haghtalab2021: two agents are said to be 
polarized if, when observing the same realization of observables, they predict the state to be different. That is, 
they both observe the same draw of $x=(x_1, ..., x_N)$ but one agent predicts the state to be **B** and the other 
predicts the state to be **R**.

Within out framework---as in @Haghtalab2021---polarization may persist even when agents have 
access to infinite data because they might be using different models of the world. 
Nevertheless using different models does not necessarily imply that beliefs will always be polarized, it must also be 
the case that the realization of observables falls in a polarization zone. By *polarization zone*, we  
mean that if the two models being used for prediction are $M_1$ and $M_2$, the realization of the 
observables $x=(x_1, ..., x_N)$ must be such that the optimal rule for $M_1$ makes a different prediction from 
that of $M_2$. The following 
2-dimensional example illustrates why the use of different models is necessary but not sufficient for polarization 
to arise.

##### Example
_The optimal rule for the complex model coincides with the truth, it predicts **B** if $x_2-1\geq x_1$. 
In addition, there are two simple models: one that considers only $x_1$---call it $M_1$---and one that considers 
only $x_2$---call it $M_2$. The optimal rules for $M_1$ and $M_2$ are threshold rules that predict the state is **B** 
if the value of $x_i$ is below (above) the threshold $t_i := \mu_{-i} + k$ and the predicted state is **R** otherwise.
These prediction rules are illustrated in . As Figure \ref{example} makes clear, both rules agree on the predicted state for 
some values of $(x_1, x_2)$ and disagree for other values. The regions in which the rules make different predictions
are the disagreement zones._

```{r example, out.width='50%', fig.align='center', fig.cap='\\label{example} Polarization regions for two single variable models'}
knitr::include_graphics(here("diagram_polariazation_2variables.png"))
```

For any pair of agents who observe the same realization of the observables and use different models, it is possible
to determine whether they will be polarized or not by looking at the predictions made by the optimal rule corresponding to their 
models. Notice that the 
polarizing observations---those that fall into a polarization zone---are different for each pair of models. Also notice that 
our definition of the polarization zone assumes that the agents are using optimal rules. This is a very strong assumption 
and will most likely not be the case that subjects in the experiment learn perfectly. This is a theoretical construct that allows
us to formalize Prediction 2.

**Prediction 2 :** Polarization will arise more often when the theory predicts so. That is, when two agents who have 
different models of the world, are facing a realization of observables that falls into the polarization zone corresponding 
to their models.

@@

Subjects for the experimet were recruited from the CESS lab subject pool at NYU. 
We recruited `r data[, uniqueN(participant.code)]` undergraduate students who participated in this experiment in person 
durinf the summer and fall of 2023. The experiment was coded using o-tree @otree and it consisted of two parts:
Part 1 was meant to expose the subjects to the data generating process
and allow them to learn how to predict; part 2 was meant to elicit the models of the world that subjects might have
developed in part 1.

Part 1 consited of 20 rounds in which subjects observed the realizations of all of the observables and were asked to predict
the state. We had 5 observable variables called variable 1, 2, 3, 4 and 5. The variables were independent and Normally 
distributed with means and variances as described in TABLE.

The state was determined by the following rule: $11x_1+6x_2+4.5x_3+2.4*x_4+k \geq 0$. The values of the coefficients were chosen
so that all variables had a similar but unique level of informativeness when considered on their own. In this case, because
all the variables are normally distributed, the information that they convey about the state is related to the
variance of the modified normal $a_ix_i$. The constant $k$ was chosen so that the state was **B** in 50% of the cases. 
The last column of TABLE shows the probability of the optimal prediction rule making a correct prediction when the model used
considers only each variable by itself. This is the measure that we use to determine the informativeness of each variable and
is reported in the last column of TABLE.

@@

In each round, subjects were asked to predict the state. They were given feedback on whether their prediction was correct or not.
Both subjects within a pair were shown exactly the same realization of the variables and had to predict the same state. This allows
us to determine whether the pair was polarized  or not for that particular observation.

Part 2 started immediatly after part 1 and has 70 rounds. In this part each subject was randomly allowed to use only a certain 
number of variables in order to make their prediction. Instead of showing them the realizations of all 5 variables, they were 
told that they can disclose up to $K$ variables, where $K$ was drawn at random from the set $\{1, 2, ..., 5\}$. K was redrawn
independently every round and across subjects. We use the random assignment of $K$ to investigate whether the subjects are using 
siumple or complex models. If allowing subjects to disclose an additional variable improves their prediction accuracy, 
then it must be that they are in fact using the additional information for prediciton. If, on the other hand, they do not perform 
better when they have access to more information, it must be that although they are disclosing an additional variable, 
they do not correctly incorporate the infomation into their prediciton rule. Thus, if we observe that the performance of our 
subjects does not improve when they are allowed to disclose $K+1$ variables, 
relative to their performance when they were allowed to disclose only $K$ variables, their model of the world must be of size $K$
or lower. This feature design allows us to test prediciton 1 and 1B and determine the number of variables that subjects are using.

In order to explore the polarization prediciton, we use the fact that subjects are in fixed pairs throughout the experiment.
Both subjects in a pair observe the same realization of all the variables in each round. Therefore, we can determine whether
they are polarized or not by looking at whether they make the same prediction or not. We can then determine whether the theory 
predicts that they should be polarized or not by looking at the polarization zone corresponding to the models that they are using.
If the draw of the variables falls into the polarization zone, then the theory predicts that they should be polarized. The models
that we use for this prediction are the ones that include the variables that they chose to reveal in that round. It could be that
although they are revealing certain variables, they are not using them for prediction, however, we do not have a good way to 
determine this. Instead, we use the variables that they reveal as a proxy for the variables that they are using for prediction and 
 determine the polarization zones for those models.

@@

### Learning
In the first part of the experiment, subjects were asked to predict the state given the realization of all the variables.
Having access to all the information they could predict the state perfectly if they managed to learn the parameters of the
hyperplane that determines the state. Figure \ref{learningAll} shows the share of correct predictions across rounds in part 1. 
We see that subjects are able to learn how to predict the state at a rate that is higher than random. They initially predict
the state corectly about 50% of the time, which is consistent with random guessing. The share of correct guesses increases to 
`r 100*round(part1[subsession.round_number>15, mean(player.correct)], digits=2)`% by the end of part 1. And this last number is 
significantly higher than 50% 
(p-value = `r round(t.test(part1[subsession.round_number>10, player.correct], mu=0.5)$p.value, digits=4)`). 
They continue to learn even after the first 20 rounds when we look at the rounds in which subjects had access to all the information.
By the end of the experiment the rate of correct predictions is 
`r 100*round(part2[subsession.round_number>30, mean(player.correct)], digits=2)`% which is still significantly higher than 50%
(p-value = `r round(t.test(part2[subsession.round_number>30, player.correct], mu=0.5)$p.value, digits=4)`).

```{r learningAll, out.width='50%', fig.align='center', fig.cap='\\label{learningAll} Share of correct predictions across rounds'}
knitr::include_graphics(here("computed_objects", "figures", "learning_all.png"))
```

>>>>

---
title: "Learning from Data Through Models"
date: '`r format(Sys.time(), "%B %d, %Y")`'
author: "Alberto Bisin, Guillaume Frechette and Jimena Galindo"
abstract: "TBW"
#classoption: pagebackref # passes pagebackref=true to hyperref before it is loaded, allowing backreferencing
output: 
    pdf_document:
        citation_package: natbib
        fig_width: 7
        fig_height: 6
        fig_caption: true
        number_sections: true
        keep_tex: true
        extra_dependencies:
            footmisc: ["bottom"] # footnote management
            setspace: ["doublespacing"] # spacing of the paper
            caption: ["normal"]
            dsfont: null # indicator function 1
            booktabs: null
            makecell: null
            hyperref: null
        includes:
          in_header: extra_header.tex

bibliography: bibliography.bib
fontsize: 12pt
geometry: margin=2.5cm
---

```{r dependencies, include=FALSE}
library(data.table)
library(here)
library(huxtable)
library(broom)
renv::snapshot()
library("languageserver")
library(reticulate)

# Format the numbers in the inline code output
inline_hook <- function(x){
  if(is.numeric(x)) {

    # integers get , to separate thousands
    if(abs(x - round(x)) < .Machine$double.eps){
      formatted <- format(x, big.mark = ",", scientific = FALSE)
    } else {
    # consistent rounding of non-integers
      formatted <- format(x, digits = 2, nsmall = 0, scientific = FALSE)
    }

  } else{

    # render text in bold (commented out)
    if(is.character(x)){
      # formatted <- paste0("**", x, "**")
      formatted <- x
    } else{
      formatted <- x
    }

  }

  formatted
}

knitr::knit_hooks$set(inline = inline_hook)
```

# Introduction

# Literature Review

@@

## Optimal Rules
Conditional of using a particular mental model, the most accurate prediction rule that the agent can use is one that 
predicts the state to be **B** whenever, given the observables under consideration, **B** is more likely. 
And it predicts **R** whenever **R** is more likely. Because we abstract away from the procedure through which the rule is updated, 
we will make predictions under the assumption that the agent has learned the parameters perfectly. That is, 
they have seen enough information that, under an updating procedure that converges to the truth---for example, 
correctly specified regression model or Bayes rule---they would have converged to the true parameters already. This provides an 
upper bound for how well any model can perform.

@@

\begin{proposition}
Fix any model that considers $m\leq N$ variables and relabel the variable so that $x_1, ..., x_m$ are the variables 
that the model considers and $x_{m+1}, ..., x_N$ are the ones that the model does not consider. Relabel the 
coefficients $a = (a_1, ..., a_N)$ accordingly as well. 
\end{proposition}
*Proof*.
Let $M:=a_1x_1+...+a_mx_m$ and $M^c = a_{m+1}x_{m+1}+...+a_Nx_N+k$ and define the latent variable $y^L := M +  M^c$. 
The optimal prediction rule for model $M$ predicts the state to be $\textbf{B}$ whenever 
$P[y^L|M]\geq \frac{1}{2}$ and predicts $\textbf{R}$ otherwise.

Using the fact that for a random variable $z\sim N(\mu_z, \sigma_z)$ we have that $P[z\geq \mu_z]=\frac{1}{2}$ and 
noticing that $M^c$ and $y^L|M$ are Normally distributed with means $\mathbb{E}[M^c]$ and $M+\mathbb{E}[M^c]$ 
respectively, it is easy to see that $P[y^L\geq 0 | M] \geq P[y^L\geq M+\mathbb{E}[M^c | M]]=\frac{1}{2}$ whenever 
$M+\mathbb{E}[M^c]\geq 0$. Similarly, whenever $M+\mathbb{E}[M^c]< 0$ we will have that 
$P[y^L\geq 0 | M] < P[y^L\geq M+\mathbb{E}[M^c | M]]=\frac{1}{2}$. Therefore, the optimal rule is to predict 
$\mathbf{B}$ whenever $a_1x_1+...+a_kx_k\geq -\mathbb{E}[M^c]$ and $\textbf{R}$ otherwise.
$\quad \square$

 \

@@

## Polarization
We take the definition of polarization from @Haghtalab2021: two agents are said to be 
polarized if, when observing the realization of X, they predict the state to be different. That is, 
they both observe the same draw of $(x_1, ..., x_N)$ but one agent predicts the state to be **B** and the other 
predicts the state to be **R**.

Within out framework---as in @Haghtalab2021---polarization may persist even when agents have 
access to unlimited data. There are two factors that dare necessary for polarization to arise: the use of different 
models and the realization of the observables must be such that the models will make different predictions. It is important
to observe that even when agents are using different models, not all realiations of $X$ will produce polarized 
predictions. The following 2-dimensional example illustrates why the use of different models is necessary but not sufficient 
for polarization to arise.

**Example.**
_The optimal rule for the complex model coincides with the truth, it predicts **B** if $x_2-1\geq x_1$. 
In addition, there are two simple models: one that considers only $x_1$---call it $M_1$---and one that considers 
only $x_2$---call it $M_2$. The optimal rules for $M_1$ and $M_2$ are threshold rules that predict the state is **B** 
if the value of $x_i$ is below (above) the threshold $t_i := \mu_{-i} + k$ and the predicted state is **R** otherwise.
These prediction rules are illustrated in . As Figure \ref{examp} makes clear, both rules agree on the predicted state for 
some values of $(x_1, x_2)$ and disagree for other values. The regions in which the rules make different predictions
are the disagreement zones._

```{r, out.width='50%', fig.align='center', fig.cap='\\label{examp} Polarization regions for two single variable models'}
knitr::include_graphics(here("diagram_polariazation_2variables.png"))
```

For any pair of agents who observe the same realization of the observables and use different models, it is possible
to determine whether they will be polarized or not by looking at the predictions made by the optimal rule corresponding to their 
models. Polarizing observations---those for which the predictions are different---are different for each pair of models. 
Also notice that we determine whether an observation is polarizing or not by using the optimal rules. We do not expect 
subjects in our experiment to converge to the optimal rule, this is a theoretical construct that allows
us to formalize Prediction 2.

**Prediction 2 :** Polarization will arise more often when the theory predicts so. That is, when two agents who have 
different models of the world, are facing a realization of observables that polarizing for their models.

@@

Subjects for the experimet were recruited from NYU's CESS-lab subject pool. 
We recruited `r data[, uniqueN(participant.code)]` undergraduate students who participated in this experiment in person 
during the summer and fall of 2023. The experiment was coded using oTree (@otree) and it consisted of two parts:
Part 1 was meant to expose the subjects to the data generating process
and allow them to learn how to predict the state; part 2 was meant to elicit the models of the world that subjects might have
developed in part 1.

Part 1 consited of 20 rounds in which subjects observed the realizations of all of the observables and were asked to predict
the state. We had 5 observable variables called variable 1, 2, 3, 4 and 5. The variables were independent and Normally 
distributed with means and variances as described in Table \ref{params}.

The state was determined by the following linear rule: $11x_1+6x_2+4.5x_3+2.4x_4+k \geq 0$. 
The values of the coefficients were chosen
so that all variables had a similar but unique level of informativeness when considered on their own. The constant $k$ was 
chosen so that the state was **B** in 50% of the cases. 
The last column of Table \ref{params} shows the probability of the optimal prediction rule making a correct prediction 
when for each single-variable model. This is the measure that we use to determine the informativeness of each variable.

@@

Table: Parameters for the Data Generating Process \label{params}

In each round, subjects were asked to predict the state. They were given feedback on whether their prediction was correct or not 
and they had access to the entire history of the game. 
Throughout the experiment, subjects were paired with another participant at random. Both subjects in a pair were shown exactly 
the same realization of the variables and had to predict the same state. This allows us to determine whether the pair was 
polarized for that particular observation. Subjects were not told that there was someone else observing the same realizations as them.

Part 2 started immediatly after part 1 and had 70 rounds\footnote{In the first 2 sessions we had 40 rounds in part 2
and the experiment was done in under 30 minutes. Because the experiment concluded in less time than expected, we increased the 
number of rounds to 70 for the following sessions. Only 12 of our subjects participated in the sessions with 40 rounds, all others 
were in sessions with 70 rounds for part 2}. 
In instead of showing subjects the realizations of all 5 variables, they were 
told that they can disclose up to $m$ variables, where $m$ was drawn at random from the set $\{1, 2, ..., 5\}$. m was redrawn
independently across rounds and across subjects.

We use the random assignment of $m$ to investigate whether the subjects are using 
siumple or complex models. If allowing subjects to disclose an additional variable improves their prediction accuracy, 
then it must be that they are in fact using the additional information for prediciton. If, on the other hand, they do not perform 
better when they have access to more information, it must be that, although they are disclosing an additional variable, 
they do not correctly incorporate the infomation into their prediction rule. Thus, if we observe that the performance of our 
subjects does not improve when they are allowed to disclose $m+1$ variables, 
relative to their performance when they were allowed to disclose only $m$ variables, their model of the world must be of size $m$
or lower. This feature design allows us to test predictions 1 and 1B and determine the number of variables that subjects are using.

In order to explore the polarization prediction, we use the fact that subjects are in fixed pairs throughout the experiment.
Both subjects in a pair observe the same realization of all the variables in each round. Therefore, we can determine whether
they are polarized or not by looking at whether they make the same prediction or not. We can then determine whether the theory 
predicts that they should be polarized or not by looking at what the optimal rules for those models dictate.
We use the optimal rules that correcpond to the variables that they chose to reveal in that round. It could be that
although they are revealing certain variables, they are not using them for prediction. Since we do not have a good method to 
determine which variables they actually pay attention to, we use the variables that they reveal as a proxy for the variables 
that they are using for prediction and apply the optimal rules for those models.

@@

\subsection{Learning}
In the first part of the experiment, subjects were asked to predict the state given the realization of all the variables.
Having access to all the information they could predict the state perfectly if they managed to learn the parameters of the
hyperplane that determines the state. Figure \ref{learningAll} shows the share of correct predictions across rounds in part 1. 
We see that subjects are able to learn how to predict the state at a rate that is higher than random. They initially predict
the state corectly about 50% of the time, which is consistent with random guessing. The share of correct guesses increases to 
`r 100*part1[subsession.round_number>15, mean(player.correct)]`% by the end of part 1. And this last number is 
significantly higher than 50% 
(p-value = `r t.test(part1[subsession.round_number>10, player.correct], mu=0.5)$p.value`). 
They continue to learn even after the first 20 rounds when we look at the rounds in which subjects had access to all the information.
By the end of the experiment the rate of correct predictions is 
`r 100*part2[subsession.round_number>30, mean(player.correct)]`% which is still significantly higher than 50%
(p-value = `r t.test(part2[subsession.round_number>30, player.correct], mu=0.5)$p.value`).

```{r learningAll, out.width='50%', fig.align='center', fig.cap='\\label{learningAll} Share of correct predictions across rounds by performance in rounds 20 to 40'}
knitr::include_graphics(here("computed_objects", "figures", "learning_all.png"))
```

We also observe that there is heterogeneity in how our subjects learn. In particular, in the first 20 rounds of part 2 we are 
able to identify two distinct groups of subjects: those who guess more than 50% of the states correctly and those who guess
50% of fewer of them correctly. As can seen in \ref{p2groups}, these two groups have very different learning patterns 
throughout the experiment. The group that does better than random in rounds 20 to 40 (the first 20 rounds of part 2) consists 
of `r part2[better_random_2040==TRUE, uniqueN(participant.code)]` subjects, which account for 
`r 100*part2[better_random_2040==TRUE, uniqueN(participant.code)]/part2[, uniqueN(participant.code)]`% of the sample. These subjects 
improved their performance from 50% to `r 100*part2[better_random_2040==TRUE & round_number_modif<41, mean(player.correct)]`% in the 
first 20 rounds of part 2, and continued to improve to get to 
`r 100*part2[better_random_2040==TRUE & round_number_modif>50 & round_number_modif<61, mean(player.correct)]`% by the end of part 2. 
Meanwhile, the group that does worse than random in rounds 20 to 40 consists of `r part2[better_random_2040==FALSE, uniqueN(participant.code)]`
subjects and they do not seem to learn how to predict the state more accurately even after all 60 rounds of the experiment.
Whenever it is relevant, we will show results separately for these two groups.

```{r p2groups, out.width='50%', fig.align='center', fig.cap='\\label{p2groups} Share of correct predictions across rounds'}
knitr::include_graphics(here("computed_objects", "figures", "p2_correct_rounds.png"))
```

## Simplicity

In part 2, subjects were allowed to disclose up to $m$ variables before making their predictions. and $m$ was assigned at random 
for every subject in every round. We use this random assignment to understand if allowing them to reveal an additional variable 
improves their prediction accuracy. If it does, it must be that they are using the additional information for prediction. 
Figure \ref{simplicity} shows the share of correct predictions by the number of variables that subjects chose to disclose.

```{r simplicity, out.width='50%', fig.align='center', fig.cap='\\label{simplicity} Share of correct predictions by number of disclosed variables'}
knitr::include_graphics(here("computed_objects", "figures", "revealed_variables_pooled.png"))
```
We see that there is a sharp increase when going from 1 to 2 variables, but no significant increase when going from 2 to 3 and
beyond. This suggests that on average subjects are using mental models that consider 2 variables. However, since subjects were 
allowed to reveal fewer variables than what was assigned to them, the estimates in \ref{simplicity} might be biased. To account 
for this, we look at the share of correct predictions by the number of variables that subjects were allowed to disclose. The results 
are presented in Table

```{r }
m1 <- lm(player.correct ~ C(player.number_variables) - 1, data = part2)
m2 <- lm(player.correct ~ C(revealed_variables_count) - 1, data = part2)

huxreg("Disclosed" = m2, 
       "Allowed" = m1,
       coefs = c("Zero variables" = "C(revealed_variables_count)0",
                 "One variable" = "C(player.number_variables)1",
                 "One variable" = "C(revealed_variables_count)1",
                 "Two variables" = "C(player.number_variables)2",
                 "Two variables" = "C(revealed_variables_count)2",
                 "Three variables" = "C(player.number_variables)3",
                 "Three variables" = "C(revealed_variables_count)3",
                 "Four variables" = "C(player.number_variables)4",
                 "Four variables" = "C(revealed_variables_count)4",
                 "Five variables" = "C(player.number_variables)5",
                 "Five variables" = "C(revealed_variables_count)5"), 
        statistics = c("N. obs." = "nobs", "R squared" = "r.squared" ))

## Model Choices

## Polarization Results
